---
title: "Data Preprocessing"
author: "Sathvik"
desription: "Data Preprocessing"
date: "10/09/2022"
format:
  html:
    df-print: paged
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - data preprocessing
---

## Libraries

```{r, include=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(dm)
library(stringr)
library(data.table)
```

## Load datasets

```{r}
duans <- read_excel("_data/n=149 with Duans.xlsx")
labworks <- read_excel("_data/091522 LABWORKS DOWNLOAD.xlsx") %>% filter(NAME == "PEARSON") %>% 
  select(-c(UNITS, STATUS))
usgs_gauge_data <- read_excel("_data/1 of 3 - USGS gage data downloaded 092022.xlsx",
                              sheet = "Pearson", guess_max = 1048576)
```


Had a problem while loading the data.`guess_max` determines how many cells in each column are used to make a guess of the column type. we can provide a `guess_max`  for read_excel to correctly guess the column type. 


## Data and the datatypes of Columns

```{r}
dm_draw(dm(duans, labworks, usgs_gauge_data), view_type = "all", column_types = TRUE, focus = TRUE)
```


cleaning columns with correct data types

labworks$RESULT is character type in the table, correcting it to numeric type.  "NA" value introduced while correcting data type in (only one)observation where result is ">24200". 

Filtered usgs_guage_data for the labworks datetime range

```{r}
labworks$RESULT <- as.numeric(labworks$RESULT)
labworks$WATERBODY <- as.factor(labworks$WATERBODY)
labworks$LCOD <- as.factor(labworks$LCOD)
labworks$PARAMETER <- as.factor(labworks$PARAMETER)
labworks <- labworks %>% rename("datetime" = "DATETIME")
# labworks <- labworks %>% rename("Name" = "NAME")

usgs_gauge_data <- usgs_gauge_data %>% 
  filter(between (datetime, min(labworks$datetime), max(labworks$datetime)))
```




```{r}
labworks <- labworks %>% 
  pivot_wider(names_from = PARAMETER, values_from =  RESULT)
  
```


Converting the dataframes to data.table and merging them to the nearest datetime.

```{r}
labworks <- data.table(labworks)
usgs_gauge_data <- data.table(usgs_gauge_data)
duans <- data.table(duans)

setkey(labworks, datetime)
setkey(usgs_gauge_data, datetime)
setkey(duans, datetime)

all_data_guage_labworks <- usgs_gauge_data[labworks , roll = "nearest" ] %>% select(-NAME)

all_data_duans <- duans[all_data_guage_labworks, roll = "nearest"]

names(all_data_guage_labworks) <- str_to_title(names(all_data_guage_labworks))
names(all_data_duans) <- str_to_title(names(all_data_duans))

```

Information about [data.table](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html) <br/>
Information about [setkey](https://www.rdocumentation.org/packages/data.table/versions/1.14.2/topics/setkey)<br/>
Information about [roll = "nearest"](https://jangorecki.gitlab.io/data.cube/library/data.table/html/data.table.html)<br/>

::: panel-tabset
## duans

```{r}
duans
```

## labworks

```{r}
labworks
```


## usgs_gauge_data

```{r}
usgs_gauge_data
```

## labworks USGS combined

```{r}
all_data_guage_labworks
```

## All data combined

```{r}
all_data_duans
```

:::

## saving merged file

```{r}
all_data <- apply(as.data.frame(all_data_guage_labworks),2 ,as.character)
write.csv(all_data, "_data/all_data_guage_labworks.csv")
```

